The curious world of constructive mathematics

### 2022-09-04 algar-lecture-notes.tex
On the double-negation embedding

@@

\begin{exercise}[Basics on negation]
Recalling that negation is defined as implying absurdity,~$\neg\varphi
\defequiv (\varphi \Rightarrow \bot)$, verify intuitionistically without
recourse to truth tables:
\begin{enumerate}
\item $\varphi \Rightarrow \neg\neg\varphi$
\item $\neg\neg\neg\varphi \Leftrightarrow \neg\varphi$
\item $\neg\neg(\varphi \vee \neg\varphi)$
\item $\neg\neg(\alpha \wedge \beta) \Leftrightarrow (\neg\neg\alpha \wedge
\neg\neg\beta)$
\item $(\neg\neg\alpha \wedge (\alpha \Rightarrow \neg\neg\beta)) \Longrightarrow
\neg\neg\beta$
\item $(\forall\psi\_ (\psi \vee \neg\psi))
\Longleftrightarrow (\forall\psi\_ ((\neg\neg\psi) \Rightarrow
\psi))$

@@

This lecture illustrates the latent constructive nature of classical
proofs with examples and presents two general metatheorems which
elucidate proof mining for constructive content.
\end{intro}

@@

\begin{proof}[Proof (classical)]The case~$k = 0$ is trivial and we omit the cases~$k
\geq 2$, hence let~$k = 1$. In this case, the map~$f$ attains some minimal
value. Set~$i$ to be (one of the) positions where this minimal value is
attained. Set~$j \defeq i+1$. Then, trivially, $f(i) \leq f(j)$.
\end{proof}

@@

What is the meaning of these non-effective proofs? Theorem~\ref{thm:dickson}
claims the existence of a finite object with a decidable property (a
pair~$(i,j)$ such that~$f(i) \leq f(j)$), but the given proof employs
transfinite methods and gives no indication how we could compute or otherwise
find this object. Instead, the classical proof asks us to grasp the infinitude
of all values of~$f$ so we can determine the minimum. The issue is even more
pronounced with Theorem~\ref{thm:surjective-matrix}, since the existence of
maximal ideals in nontrivial rings requires the axiom of
choice~\cite{scott:prime-ideals,hodges:krull,banaschewski:krull,erne:krull,howard-rubin:ac}.
To add one more conundrum: Assume that we have a classical proof, using the principle
of excluded middle and the axiom of choice, that some given Turing machine
terminates. Can we then constructively accept that the machine will halt? Do we
have an upper bound for the number of computational steps the machine carries
out before halting?

@@

\begin{enumerate}
\item \emph{Eliminating the axiom of choice by the~$L$-translation.} Can the
axiom of choice ever help in proving arithmetical statements, those first-order
statements in which all quantifiers range over the natural numbers? Well, it
might. But a result of Gödel states that~\textsc{zfc} (Zermelo--Fraenkel set
theory with the axiom of choice) is conservative over~\textsc{zf} (ZF set
theory without it) for such statements -- hence all appeals to this axiom can
be mechanically eliminated from a given proof. This is true even if the proof
transcends the arithmetical realm and includes statements which are not
arithmetical, only the asserted claim is required to be arithmetical. Hence we
are free to use the axiom of choice, tranquil in knowing that we could always
reformulate our proofs without it.\footnote{Zermelo--Fraenkel set theory with
the axiom of choice is the go-to foundation of mathematics often cited as
supporting ``almost all'' of current mathematics, one important exception being
some (definitions and) results in category theory dealing with large
structures~\cite{shulman:set-theory,feferman:set-foundations}. A fundamental
result due to Gödel is that the axiom of choice ``holds in~$L$'', the
\emph{constructible universe}, even if it might not hold in~$V$, the true
universe of all sets. More precisely, if~\textsc{zfc} shows some
statement~$\varphi$, then~\textsc{zf} shows its~$L$-relativized
version~$\varphi^L$, where all quantifiers have been restricted to range
over~$L$ instead of~$V$. The conservation result follows because the natural
numbers ``are absolute between~$V$
and~$L$''~\cite{goedel:ac-gch,schoenfield:predicativity}.}

\item \emph{Eliminating the principle of excluded middle by the double-negation
translation and its variants.} At the price of
slightly modifying the asserted claim, the principle of excluded middle
can always be mechanically eliminated from a given proof. This
elimination procedure is facilitated by the \emph{double-negation translation}
reviewed below. In some cases, a refined translation even allows to preserve the asserted
claim exactly; this technique is variously known as \emph{Friedman's trick},
\emph{nontrivial exit continuation} or (the baby version of) \emph{Barr's
theorem}. To cite a concrete instance of this phenomenon, classical~\textsc{zf}
set theory is conservative over its intuitionistic cousin~\textsc{izf}
for~$\Pi^0_2$-statements (statements of the form~$\forall\ldots\forall\_
\exists\ldots\exists\_ \%$, where all quantifiers in the final~``$\%$''
are bounded).

\item \emph{Embracing generic models.} A useful companion to both of the
aforementioned techniques is to switch from referencing all models of a certain
kind to referencing only the \emph{generic model}. For instance, Krull's lemma
stating that a ring element is already nilpotent if it is contained in all
prime ideals requires the Boolean Prime Ideal Theorem, a slightly weaker
version of the axiom of choice but still ineffectiv and unconstructive.
However, Krull's lemma is valid in the form that a ring element is nilpotent if
it is contained in the \emph{generic prime ideal}. This particular example has
received lots of attention (see the references
in~\cite{blechschmidt-schuster:constructive-maximal-ideals}) and the general
technique will be the object of the fourth lecture.
\end{enumerate}

Noticeably missing in this list is any technique for eliminating uses of the
\emph{powerset axiom} stating that the collection of all subsets of a given
set is again a set. While this axiom is uncontested by ordinary constructive
mathematics and doesn't receive nearly as much philosophical attention as the
axiom of choice or the principle of excluded middle, it is this axiom which actually
and substantially increases logical strength. While~\textsc{zfc}, \textsc{zf}
and \textsc{izf} are equiconsistent (and in fact verify the same
arithmetical~$\Pi^0_2$-statements), systems without it such as Kripke--Platek
set theory~(\textsc{kp}) or constructive Zermelo--Fraenkel set
theory~(\textsc{czf}) are much weaker. We will not discuss this curious state
of affairs; references include~\cite{crosilla:predicativity,aczel-rathjen:cst}.

\section{The double-negation translation}
\newpage

@@

\begin{exercise}[Details on the double-negation translation]
Complete the verification of the fundamental properties of the double-negation
translation.
\end{exercise}

@@

Cite surveys of .bib

>>>>

\newcommand{\cl}{\mathrm{cl}}
\newcommand{\notnot}{\emph{not~not}}

@@

\begin{exercise}[Basics on negation]
Recalling that negation is defined as implying absurdity,~$\neg\varphi
\defequiv (\varphi \Rightarrow \bot)$, verify intuitionistically without
recourse to truth tables:
\begin{enumerate}
\item $\varphi \Rightarrow \neg\neg\varphi$
\item $\neg\neg\neg\varphi \Leftrightarrow \neg\varphi$
\item $\neg\neg(\varphi \vee \neg\varphi)$
\item $\neg\neg(\alpha \wedge \beta) \Leftrightarrow (\neg\neg\alpha \wedge
\neg\neg\beta)$
\item $\neg(\alpha \vee \beta) \Longleftrightarrow (\neg\alpha \wedge \neg\beta)$
\item $(\neg\neg\alpha \wedge (\alpha \Rightarrow \neg\neg\beta)) \Longrightarrow
\neg\neg\beta$
\item $(\forall\psi\_ (\psi \vee \neg\psi))
\Longleftrightarrow (\forall\psi\_ ((\neg\neg\psi) \Rightarrow
\psi))$

@@

This lecture illustrates the latent constructive nature of classical
proofs with examples and presents two general metatheorems which
elucidate proof mining for constructive content. The author learned this
material from diverse texts and slides of Thierry Coquand; the
reference~\cite{coquand:classical} is a good start.
\end{intro}

@@

\begin{proof}[Proof (classical)]The case~$k = 0$ is trivial and we omit demonstrations of the cases~$k
\geq 2$, hence let~$k = 1$. In this case, the map~$f$ attains some minimal
value. Set~$i$ to be (one of the) positions where this minimal value is
attained. Set~$j \defeq i+1$. Then, trivially, $f(i) \leq f(j)$.
\end{proof}

@@

What is the meaning of these non-effective proofs? Theorem~\ref{thm:dickson}
claims the existence of a finite object with a decidable property (a
pair~$(i,j)$ such that~$f(i) \leq f(j)$), but the given proof employs
transfinite methods and gives no indication how we could compute or otherwise
find this object. Instead, the classical proof asks us to grasp the infinitude
of all values of~$f$ and determine their minimum. The issue is even more
pronounced with Theorem~\ref{thm:surjective-matrix}, since the existence of
maximal ideals in nontrivial rings requires the axiom of
choice~\cite{scott:prime-ideals,hodges:krull,banaschewski:krull,erne:krull,howard-rubin:ac}.
To add one more conundrum: Assume that we have a classical proof, using the principle
of excluded middle and the axiom of choice, that some given Turing machine
terminates. Can we then constructively accept that the machine will halt? Do we
have an upper bound for the number of computational steps the machine carries
out before halting?

@@

\begin{enumerate}
\item \emph{Eliminating the axiom of choice by the~$L$-translation.} Can the
axiom of choice ever help in proving arithmetical statements, those first-order
statements in which all quantifiers range over the natural numbers? Well, it
might. But a result of Gödel states that~\textsc{zfc} (Zermelo--Fraenkel set
theory with the axiom of choice) is conservative over~\textsc{zf} (ZF set
theory without it) for such statements -- hence all appeals to this axiom can
be mechanically eliminated from a given proof. This is true even if the proof
transcends the arithmetical realm and includes statements which are not
arithmetical; only the asserted claim is required to be arithmetical. Hence we
are free to use the axiom of choice, tranquil in knowing that we could always
reformulate our proofs without it.\footnote{Zermelo--Fraenkel set theory with
the axiom of choice is the go-to foundation of mathematics often cited as
supporting ``almost all'' of current mathematics, one important exception being
some (definitions and) results in category theory dealing with large
structures~\cite{shulman:set-theory,feferman:set-foundations}. A fundamental
result due to Gödel is that the axiom of choice ``holds in~$L$'', the
\emph{constructible universe}, even if it might not hold in~$V$, the true
universe of all sets. More precisely, if~\textsc{zfc} shows some
statement~$\varphi$, then~\textsc{zf} shows its~$L$-relativized
version~$\varphi^L$, where all quantifiers have been restricted to range
over~$L$ instead of~$V$. The conservation result follows because the natural
numbers ``are absolute between~$V$
and~$L$''~\cite{goedel:ac-gch,schoenfield:predicativity}.}

\item \emph{Eliminating the principle of excluded middle by the double-negation
translation and its variants.} At the price of
slightly modifying the asserted claim, the principle of excluded middle
can always be mechanically eliminated from a given proof. This
elimination procedure is facilitated by the \emph{double-negation translation}
reviewed below. In some cases, a refined translation even allows to preserve the asserted
claim exactly; this technique is variously known as \emph{Friedman's trick},
\emph{nontrivial exit continuation} or (the baby version of) \emph{Barr's
theorem}. To cite a specific instance of this phenomenon, classical~\textsc{zf}
set theory is conservative over its intuitionistic cousin~\textsc{izf}
for~$\Pi^0_2$-statements (statements of the form~$\forall\ldots\forall\_
\exists\ldots\exists\_ \%$, where all quantifiers in the final~``$\%$''
are bounded).

\item \emph{Embracing generic models.} A useful companion to both of the
aforementioned techniques is to switch from referencing all models of a certain
kind to referencing only the \emph{generic model}. For instance, Krull's lemma
stating that a ring element is already nilpotent if it is contained in all
prime ideals requires the Boolean Prime Ideal Theorem, a slightly weaker
version of the axiom of choice but still ineffective and unconstructive.
However, Krull's lemma is valid in the form that a ring element is nilpotent if
it is contained in the \emph{generic prime ideal}. This particular example has
received lots of attention (see the references
in~\cite{blechschmidt-schuster:constructive-maximal-ideals}) and the general
technique will be the object of the fourth lecture.
\end{enumerate}

Noticeably missing in this list is any technique for eliminating uses of the
\emph{powerset axiom} stating that the collection of all subsets of a given
set is again a set. While this axiom is uncontested by ordinary constructive
mathematics and doesn't receive nearly as much philosophical attention as the
axiom of choice or the principle of excluded middle, it is this axiom which actually
and substantially increases logical strength. While~\textsc{zfc}, \textsc{zf}
and \textsc{izf} are equiconsistent (and in fact verify the same
arithmetical~$\Pi^0_2$-statements), systems without the powerset axiom such as Kripke--Platek
set theory~(\textsc{kp}) or constructive Zermelo--Fraenkel set
theory~(\textsc{czf}) are much weaker. We will not discuss this curious state
of affairs and only note that rejecting the powerset axiom can be
well-motivated and is the basis of \emph{predicative mathematics};
references include~\cite{crosilla:predicativity,aczel-rathjen:cst}.

\section{The double-negation embedding of classical into constructive mathematics}

The premier difference between constructive and classical mathematics is in the
existential quantifier~``$\exists$'' and disjunction~``$\vee$''.
Constructively, to verify an existential statement, it is not enough to verify
the impossibility of nonexistence; and to verify a disjunction~$\alpha \vee
\beta$, it is not enough to verify the impossibility of~$\neg\alpha \wedge
\neg\beta$.

That is not to say, however, that the more informative versions of~``$\exists$''
and~``$\vee$'' of constructive mathematics would be in any senser ``better''
than their classical counterparts: There are many situations in which the
classical semantics is exactly the appropriate one. Should we somehow combine
classical and intuitionistic logic to form a joint logic supporting
both the informative connectives from intuitionistic logic and the ``platonic''
connectives from classical mathematics?

Perhaps surprisingly, it turns out that no such combination is necessary, for
the classical connectives are already definable in intuitionistic logic:
\begin{align*}
  \exists^\cl x \? X\_ \varphi(x)
    &\quad\defequiv\quad \neg\neg (\exists x\?X\_ \varphi(x))
    && \text{(this is eqv. to $\neg(\forall x\?X\_ \neg\varphi(x))$)} \\
  \alpha \vee^\cl \beta
    &\quad\defequiv\quad \neg\neg (\alpha \vee \beta)
    && \text{(this is eqv. to $\neg(\neg\alpha \wedge \neg\beta)$)}
\end{align*}

This observation is the starting point of the \emph{double-negation embedding}
of classical logic into intuitionistic logic. This embedding translates any
statement~$\varphi$ into a related statement~$\varphi^{\neg\neg}$, substituting
all occurences of~``$\exists$'' and~``$\vee$'' by~``$\exists^\cl$''
and~``$\vee^\cl$'' (and prefixing all atomic statements by~``$\neg\neg$'')
as detailed in Table~\ref{table:negneg}. Crucially, while the principle of
excluded middle cannot be verified for~``$\vee$'', the principle of excluded
middle for~``$\vee^\cl$'' is an intuitionistic tautology:

\begin{thm}For every statement~$\varphi$,
$\neg\neg(\varphi \vee \neg\varphi)$.
\end{thm}

\begin{proof}By definition of negation, we are to prove
\[ \underbrace{\bigl((\varphi \vee (\varphi\Rightarrow\bot)) \Longrightarrow
\bot\bigr)}_{\equiv\vcentcolon\, \chi}
\Longrightarrow \bot. \]
So assume~$\chi$; we are to show~$\bot$.

Hypothetically, if~$\varphi$, then in particular~$\varphi \vee
(\varphi\Rightarrow\bot)$ and hence, by~$\chi$, $\bot$. This argument
establishes~$\varphi \Rightarrow \bot$.

Thus in particular we have~$\varphi \vee (\varphi \Rightarrow \bot)$ and hence,
by~$\chi$ again, $\bot$.
\end{proof}

The fundamental properties of the double-negation translation are summarized by
the following theorem.

\begin{thm}\label{thm:negneg}
For every statement~$\varphi$, \ldots
\begin{enumerate}
\item classical logic proves~$\varphi \Leftrightarrow \varphi^{\neg\neg}$,
\item intuitionistic logic proves~$\neg\neg(\varphi^{\neg\neg}) \Rightarrow
\varphi^{\neg\neg}$,
\item (if~$\varphi$ is a geometric formula) intuitionistic logic
proves~$\varphi^{\neg\neg} \Leftrightarrow \neg\neg\varphi$, and
\item classical logic proves~$\varphi$ from some set~$\Gamma$ of assumptions
if and only if intuitionistic logic proves~$\varphi^{\neg\neg}$ from the
set~$\Gamma^{\neg\neg}$ of translated assumptions.
\end{enumerate}
\end{thm}

\begin{proof}Instructive exercise.\end{proof}

The double-negation embedding is not only conceptually pleasing, establishing
that intuitionistic logic can serve as a common home for both classical and
constructive mathematics, but also supplies us with a vast source of
constructive results.

\begin{ex}Classical linear algebra teaches us that every finite system of
generators of a vector space over a residue field\footnote{Constructively, the
notion of a field bifurcates into several nonequivalent notions: A nontrivial
commutative ring with unit is a \ldots
\begin{enumerate}
\item a \emph{geometric field} or \emph{discrete field} iff
every element is zero or
invertible (as in~$\QQ$ or~$\overline{\QQ}$),
\item a \emph{residue field} iff
every element which is not invertible is zero (as in the Cauchy or Dedekind
reals), and
\item a \emph{field of fractions} iff
every element which is not zero is invertible (as in the localization of the
Cauchy or Dedekind reals at the set of non-zeros).
\end{enumerate}These each have their
uses~\cite{johnstone:rings-fields-and-spectra,mulvey:repr}.}
can be thinned to a generating family in which no vector is redundant, hence to
a basis. By slightly massaging the result obtained by the double-negation
embedding, we obtain the constructive theorem that every finitely generated
vector space over a residue field does \emph{not not} possess a basis.

We will learn in Section~\ref{sect:generic-freeness} that this simple theorem
immediately entails (a basic version of) \emph{Grothendieck's generic freeness
lemma}, a result in algebraic geometry fundamental to the theory of moduli
spaces, if applied after localizing at the ``generic prime filter'' (this
requires passing to a more adapted topos). It is marvelous how large the impact
of this simple theorem from undergraduate linear algebra is, if made to apply
in arbitrary toposes by the double-negation embedding and then specializing to
a particularly well-adapted one.\end{ex}

@@

\begin{exercise}[Details on the double-negation translation]
Complete the verification of the fundamental properties of the double-negation
translation stated as Theorem~\ref{thm:negneg}.
\end{exercise}

\begin{exercise}[Minima of sets of natural numbers]
In Theorem~\ref{thm:min}(b) we showed that every inhabited set of natural
numbers does \notnot contain a minimal element. Explain how this result follows
from the double-negation embedding applied to the classical fact that every
inhabited set of natural numbers directly contains a minimal element.

{\noindent\scriptsize\emph{Hint.} It is not enough to just cite the
double-negation embedding. The resulting constructive theorem has to be
massaged a bit to yield Theorem~\ref{thm:min}(b).\par}
\end{exercise}

\begin{exercise}[Grothendieck's generic freeness lemma, part one]
XXX
\end{exercise}

@@

\[
  \forall x_1,\ldots,x_n\?V\_
    \bigl(\forall x\?V\_ \exists a_1,\ldots,a_n\?K\_ x = \sum_i a_i x_i\bigr)
      \Longrightarrow
        \]
